---
title: "CC Challenge 2!"
author: "Edwin Yánez"
date: "2025-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
```

## Purpose
In this document I showcase my skills learned during the second course on CC. The instructions are taken as they appear on their website at [Wiz of Data Vis challenge: Native woodlands](https://ourcodingclub.github.io/course/wiz-viz-challenge/index.html). As one is encouraged to use the provided version of the data, I have also avoided including the links to the original sources here; they are still available over on the Challenge 2 page.

## Data Overview
### Native Woodland Survey of Scotland
+ `NWSS.RDS`: a shapefile of all woodland patches in Scotland. The most important variables in the dataset are:
  + `DOM_HABITA`: the main habitat type for the polygon. We will only retain some habitats of interest.
  + `HECTARES`: the area of a given patch

+ `species_structure.RDS`: a spreadsheet containing tree species information from the woodlands. The most important variables in the dataset are:
  + `SCPTDATA_I`: a unique identifier code that will allow to match the observations to the spatial data in NWSS.RDS
  + `SPECIES`: the name of the species recorded
  + `ESTIMT_HA`: the estimated area, in hectares, covered by a given species at this location

### National Nature Reserves
+ `SNH_national_reserves.RDS`: a shapefile containing the outlines of Scotland’s NNRs. The most important variables in the dataset are:
  + `NAME`: The name of the reserve
  + `SITE_HA`: The area of the site in hectares

## About spatial data
Two of the three datasets are **shapefiles**, which means that they contain geometric information that allow the data to be represented as shapes (polygons), points or lines. But don’t panic! When you import the files into R, you will see that you can preview and manipulate the data much like any other dataframe.

The spatial objects have been saved using the *sf package* which allows for integration with the tidyverse: the sf functions are pipe-friendly and you can pretty much do everything to a sf object that you would do to a regular dataframe (e.g. merge with another dataset, subset to some values or conditions, etc). Remember, in the end, a spatial dataset is just like any other dataset, with extra geographic information tucked in one column!

You will not have to do any complex spatial analysis for this, but the instructions will point you in the right direction when functions specific to the sf package might be needed.

## Specific Tasks
Here is a detailed list of the tasks you should achieve within this challenge. Remember that a challenge is meant to be, well, challenging, and therefore we are setting you goals but the choice of workflow and functions to achieve them is up to you! We also list the questions that will be asked in the quiz at the end to confirm your successful completion - we suggest you take note of your answers as you go.

### 1. Clean the data
You will need to clean and filter the data to the sites and woodland types of interest. *Specifically, you should*:

+ Restrict the NWSS observations to the following dominant habitat types:
	+ Native pinewood
  + Upland birchwood
  + Upland mixed ashwood
  + Upland oakwood
  + Wet woodland
  + Lowland mixed deciduous woodland
	
+ Restrict the NNR shapefile to the **following areas**, lump the last three under the same name, and rename as indicated:
  + The Great Trossachs Forest (rename to “Trossachs”)
  + Glen Affric (leave as such)
  + Cairngorms (part of the “Cairngorms” group)
  + Mar Lodge Estate (part of the “Cairngorms” group)
  + Abernethy (part of the “Cairngorms” group)

NB: There are 6 more NNRs within the Cairngorms National Park, but these three are large ones within the core of the park, and the only ones we’ll be considering for this analysis.

**HINT:** Once you have filtered both datasets to only keep the regions and habitats of interest, the best way forward is to create **one object** that combines the two: i.e. only keep the habitats of interest that are found within the regions of interest. You may need some indepent research to figure it out, but only one function from the sf package is required to achieve this. To get you started, know that all sf functions begin with “st“, and this type of spatial operation is called an _intersection…
```{r}
NWSS_raw <- readRDS('data/NWSS.RDS')
species_structure_raw <- readRDS('data/species_structure.RDS')
SNH_national_reserves_raw <- readRDS('data/SNH_national_reserves.RDS')

# Reassigning them to update the old-style crs:
st_crs(NWSS_raw) <- st_crs(NWSS_raw)
st_crs(SNH_national_reserves_raw) <- st_crs(SNH_national_reserves_raw)

nwss <- NWSS_raw %>% 
  filter(DOM_HABITA %in% c('Native pinewood', 'Upland birchwood', 'Upland mixed ashwood', 'Upland oakwood', 'Wet woodland', 'Lowland mixed deciduous woodland')) %>% 
  drop_na()

snh <- SNH_national_reserves_raw %>% 
  filter(NAME %in% c('The Great Trossachs Forest', 'Glen Affric', 'Cairngorms', 'Mar Lodge Estate', 'Abernethy')) %>%
  mutate(
    NAME = case_when(
      NAME == 'The Great Trossachs Forest' ~ 'Trossachs',
      NAME %in% c('Mar Lodge Estate', 'Abernethy') ~ 'Cairngorms',
      TRUE ~ NAME
    )
  ) %>% 
  drop_na()
```







































