---
title: "CC Spatial Master follow along"
author: "Edwin Yánez"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# It seems that the instructions on the tutorila are way out of date so these are new libraries I'll be using to to accomplish roughly the same tasks.
library(terra)   # For raster data operations
library(sf)      # For vector data operations
library(ggplot2) # For advanced plotting
library(viridis) # For color palettes
library(tmap)    # For quick thematic mapping
library(reshape2)
```

## Purpose
This is a follow-along document reporting my engagement with Coding Club's [Intro to spatial analysis in R](https://ourcodingclub.github.io/tutorials/spatial/index.html) tutorial. The tutorial seems severely out of date. Some parts are represented here while others were updated using an LLM tool to update all the content so that I still get to know common practices in spatial analysis with R while also learning the newest best practice with R as a whole.

## Explore raster data

A raster is a grid of equal size cells, or pixels in satellite images, and it is commonly used to represent spatially continuous data. The cells can have one or more values, or even no values for the variable of interest. In the trimmed multispectral image we will be using, each cell contains relfectance data for 12 spectral bands.

```{r}
# Load data
tay <- rast('taycrop.tif')

print(tay)
```

The output will include the number of bands, dimensions, resolution, extent, and CRS (Coordinate Reference System). Notice that—even though the original file may contain 12 bands—terra stores all bands in one SpatRaster object.

We can create individual raster layers for each of the spectral bands in the raster tay.

To work on individual spectral bands, you can extract them directly from the multi-band object. With terra you use the double-bracket operator [[ ]]:

```{r}
b1 <- tay[[1]]
b2 <- tay[[2]]
b3 <- tay[[3]]
b4 <- tay[[4]]
b5 <- tay[[5]]
b6 <- tay[[6]]
b7 <- tay[[7]]
b8 <- tay[[8]]
b9 <- tay[[9]]
b10 <- tay[[10]]
b11 <- tay[[11]]
b12 <- tay[[12]]
```

We can now compare two bands to see if they have the same extent, number of rows and column, projection, resolution and origin. As can be seen below, bands 2 and 3 match.

If you wish to compare geometries (extent, resolution, CRS) of two bands, use the `compareGeom()` function. Should return TRUE if they are identical

```{r}
compareGeom(b2, b3)
```

Checking the coordinate systems and extents of rasters is a very useful skill - quite often when you have problems with working with multiple raster objects, it’s because of differences in coordinate systems or extents.

The default plot() function in terra displays a subsample (about 100,000 pixels) to keep plots fast, while image() will render a stretched full-resolution view.

```{r}
plot(b8)   # Quick plot with a pixel limit
image(b8)   # Full-resolution plot
```

## Visualise spectral bands
You can save a plot of a specific band (e.g., b8) using the png() function. Note that the viridis package now provides its palette via the function `viridis::viridis()`, which is preferred over older functions.
```{r}
# Save a plot of band 8 with a viridis palette
png('01tayplot.png', width = 4, height = 4, units = "in", res = 300)
image(b8, col = viridis::viridis(10, option = "D"), 
      main = "Sentinel 2 image of Loch Tay")
dev.off()   # Closes the device, ensuring no conflicts with subsequent plots
```


To view the plot interactively (without saving), simply run:
```{r}
image(
  b8, 
  col = viridis::viridis(10, option = "D"),
  main = "Sentinel 2 image of Loch Tay"
  )
```

A useful way to visualise the satellite data is to plot a red-green-blue plot of a multi-layered object for a more realistic rendition. The layers or bands represent different bandwidth in the visible electromagnetic spectrum (corresponding to red, blue and green) and combined, create a naturalistic colour rendition of the earth surface.

A false-color composite (FCC) can help accentuate vegetation. For Sentinel-2, a common FCC is to replace the red band with the near infrared band (b8), use red (b4) for green, and green (b3) for blue:
```{r}
# Save a false-color composite image
png('02FCC.png', width = 5, height = 4, units = "in", res = 300)
tayFCC <- c(b8, b4, b3)  # FCC: NIR, Red, Green
plotRGB(tayFCC, 
        axes = TRUE, 
        stretch = "lin", 
        main = "Sentinel False Color Composite (FCC)"
        )
dev.off()

```

For more flexible visualizations, you can convert a single band (e.g., b8) into a data frame and plot it using ggplot2:
```{r}
library(ggplot2)

# Convert band 8 to a data frame with x, y coordinates and pixel values. This function is part of terra.
df_b8 <- as.data.frame(b8, xy = TRUE)

# Create the plot using ggplot2
p <- ggplot(df_b8, aes(x = x, y = y, fill = taycrop_8)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_quickmap() +
  labs(
    title = "West of Loch Tay, Raster Plot",
    x = "Longitude",
    y = "Latitude"
  )+
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 20),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
print(p)
ggsave("03ggtay.png", p, scale = 1.5, dpi = 300)
```

Note that here we saved the plot in a slightly different way - for plots creates using `ggplot2`, we can use the `ggsave` function and we define the specifics of the saved plot after we’ve created it, whereas earlier in the tutorial when we were using the `png()` function in combination with `dev.off()`, the plot characteristics are defined before we make the plot inside the `png()` function.

To visualise all the bands together, we can use facet_wrap in gplot. First, we will create a stack of all the bands, so just putting them all on top of each other, like layers in a cake.

```{r}
# Stack all bands using terra's c() function
t <- c(b1, b2, b3, b4, b5, b6, b7, b8, b9, b10, b11, b12)

# Convert the multi-layer raster to a data frame including an identifier for each band
t_df <- as.data.frame(t, xy = TRUE)
t_df_melt <- melt(t_df, id.vars = c("x", "y"))

# Create faceted plots for each band
p_all <- ggplot(t_df_melt, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c() +
  facet_wrap(~variable) +
  coord_quickmap() +
  labs(title = "Sentinel 2 Loch Tay, Raster Plots",
       x = "Longitude", y = "Latitude") +
  theme_classic() +
  theme(text = element_text(size = 20),
        axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5))
print(p_all)
ggsave("04allbands.png", p_all, scale = 1.5, dpi = 300)
```

Terra is designed as a modern replacement for the raster package, and it provides its own, very efficient plotting functions for SpatRaster objects. For example, if you have a multi-band raster file, you can quickly visualize it using terra’s built‐in plot methods without converting it into a brick first. Here’s how you can do it:

```{r}
# Load the multi-band raster
s_tay <- rast("taycrop.tif")

# Quickly plot all bands
plot(s_tay)

# For an RGB composite, you can use plotRGB
# (adjust the band numbers as appropriate for your data)

# Use the stretch argument so that terra rescales your band values for on‐the‐fly display. For example:
plotRGB(s_tay, r = 4, g = 3, b = 2, stretch = "lin")

# Or
plotRGB(s_tay, r = 4, g = 3, b = 2, stretch = "hist")
```

Notice the difference in colour and range of legend between the different bands. Different earth surfaces reflect the solar radiation differently and each raster layer represents how much incident solar radiation is reflected at a particular wavelength bandwidth. Bands 6 to 9 are in the Near Infrared Range (NIR). Vegetation reflects more NIR than other wavelengths but water absorbs NIR, therefore the lighter areas with high reflectance values are likely to be vegetation and the dark blue, low reflectance value areas, likely to be water. Also note that the Sentinel 2 bands have 3 levels of spatial resolution, 10 m, 20 m, and 60 m (see summary below).

10 m resolution band 2, band 3, band 4 and band 8

20 m resolution band 5, band 6, band 7, band 11 and band 12

60 m resolution band 1, band 9 and band 10

































