---
title: "CC-4-DVM-Follow-along"
author: "Edwin Yánez"
date: "2025-02-26"
output: html_document
---

```{r setup, include=FALSE}
# Needed libraries
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)
library(gridExtra)
library(vroom)

knitr::opts_chunk$set(echo = TRUE)
```

## Purpose.

This is a follow-along document reporting my engagement with Coding Club's [Beautiful and informative data visualisation](https://ourcodingclub.github.io/tutorials/datavis/index.html) tutorial. Most of the content is lifted directly from there with minor modifications and updates from my part. Not everything in the tutorial should be expected to be replicated here.

## Making different plots with ggplo2

The following uses data from the [Living Planet Index](https://www.livingplanetindex.org/).

```{r}
# Load the data
LPI <- vroom::vroom('LPIdata_CC.csv')
LPI <- LPI %>% 
  mutate(`1970` = as.numeric(`1970`))

# Reshape from wide to long
LPI2 <- LPI %>% 
  pivot_longer(
    cols = 9:53,
    names_to = 'year',
    values_to = 'abundance'
  )
```


# 