---
title: "CC Model Design Master"
author: "Edwin Yánez"
date: "2025-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(vroom)
library(lme4)
library(sjPlot)   # Probably outdated. Look into tidybayes and ggdist for Bayesian workflows
library(ggeffects)   # Probably outdated. Look into tidybayes and ggdist for Bayesian workflows
# modelbased and performance are modern ways to visualize uncertainty and model diagnostics
library(MCMCglmm)
library(MCMCvis)
# brms (which uses Stan under the hood) or rstanarm are now preferred for Bayesian hierarchical modeling due to their flexibility, active development, and better integration with the tidyverse ecosystem.
library(stargazer)
```

## Purpose
This is a follow-along document reporting my engagement with Coding Club's [Intro to model design
](https://ourcodingclub.github.io/tutorials/model-design/index.html) tutorial. Not everything in the tutorial should be expected to be replicated here.

## 1. Learn what a statistical model is
In order to answer research questions, we require statistical tests of the relationships in our data. In modern ecology and other fields, models are designed to fit the structure of data and to appropriately test the questions that we have as researchers. Thus, the first step to any data analysis is figuring out what your research question is. **Without a research question, there is no point in trying to conduct a statistical test. So let’s pause here and figure out the research question for our tutorial today**.

## 2. The research question
In this tutorial, we will work with part of the long-term plant cover dataset from the Toolik Lake Field Station. These data (remember the word data is plural, thus data are … not data is …!) are plant composition data collected over four years across five sites over time in Arctic tundra in Northern Alaska. A simple question we might ask with these data is: how has the species richness changed in these plots over time?

### Question 1: How has plant species richness changed over time at Toolik Lake?
Once we have figured out our research question, we next need to figure out our hypothesis. To come up with a hypothesis, we need to learn something about this system. To start us off today, we will suggest a hypothesis for you: plant species richness is increasing over time. We might expect this as these tundra plots might be undergoing warming and warming might lead to increased plant species richness in tundra plant communities (see [this paper](https://doi.org/10.1038/s41586-018-0005-6) for more on this topic).

**Hypothesis 1: Plant species richness has increased over time at Toolik Lake**. (Remember to phrase your hypothesis in the past tense, as these data represent changes that have already occurred and remember that results are always written in the past tense.)

Now that we have a hypothesis, it is good practice to also write a null hypothesis. What are the hypotheses that we are testing between here? For example, a null hypothesis for these data and this question might be:

**Null Hypothesis: Plant species richness has not changed over time at Toolik Lake**.

We might also have an alternative hypothesis:

**Hypothesis 2: Plant species richness has decreased over time at Toolik Lake**.

Toolik Lake Station is in Alaska, a place that has been warming at rates higher than the rest of the world, so we might also wonder how temperature influences the plant communities there, in particular, their richness. So, we pose a second question:

### Question 2: How does mean annual temperature influence plant species richness?
**Hypothesis 1: Higher temperatures correspond with higher species richness**.

How are questions 1 and 2 different?

**Detection models** When we ask how plant species richness has changed over time, we are interested in **detecting** change. We want to know what happened to plant communities in Toolik Lake, but we are not testing anything regarding **why** such changes in species richness occurred (and maybe there were no changes over time).

**Attribution models** When we ask how temperature influences plant species richness, we are looking to attribute the changes we’ve seen to a specific driver, in this case, temperature. Attribution models often are the next step from a detection model. First, you want to know what happened, then you try to figure out why it happened. For example, if we find a strong positive relationship between temperature and species richness (e.g., as temperature goes up, so does species richness), then temperature is likely to be one of the drivers of local-scale changes in species richness.

For now, this should be enough set-up for us to progress ahead with our models, but remember to **always start with the question first when conducting any research project and statistical analysis**.

## 3. Thinking about our data
There are different statistical tests that we could use to conduct our analyses and what sort of statistical test we use depends on the question and the type of data that we have to test our research question. Since we have already thought about our question for a bit, let’s now think about our data. What kind of data are we dealing with here?

Our data consists of plant species cover measured across four years in plots that were within blocks, which were then within sites. We have the variables: `Year`, `Site`, `Treatment`, `Block`, `Plot`, `Species`, `Relative.Cover`, `Mean.Temp` and `SD.Temp` in our dataframe. Let’s look at the dataframe now.
```{r}
toolik_plants <- vroom::vroom('toolik_plants.csv')

str(toolik_plants)
```

Site and Species are of the character type (text, composed of letters) - they are names, and we will treat them as categorical variables. Year, Cover, Mean.Temp and SD.Temp are numeric and continuous data - they are numbers. Cover shows the relative cover (out of 1) for different plant species, Mean.Temp is the mean annual temperature at Toolik Lake Station and SD.Temp is the standard deviation of the mean annual temperature. Then, we have Treatment, another categorical variable that refers to different chemical treatments, e.g. some plots received extra nitrogen, others extra phosphorus. Finally, we have Block and Plot, which give more detailed information about where the measurements were taken.

The plot numbers are currently coded as numbers (num) - 1, 2,…8, making it a numerical variable. We should make them a categorical variable, since just like Site and Block, the numbers represent the different categories, not actual count data.

In R, we can use the factor type to denote a vector/column as categorical data. With the following code, we can convert multiple columns to factors:
```{r}
toolik_plants <-
  toolik_plants %>% 
  mutate(
    across(c(Site, Block, Plot), as.factor)
  )
str(toolik_plants)
```

Now, let’s think about the distributions of the data. **Our data structure is a bit like a Russian doll, so let’s start looking into that layer by layer**.
```{r}
unique(toolik_plants$Site)
length(unique(toolik_plants$Site))
```

First, we have five sites (06MAT, DH, MAT, MNT and SAG).
```{r}
# Group by Site to see the number of blocks per sie
toolik_plants %>%
  group_by(Site) %>% 
  summarise(
    block.n = n_distinct(Block),
    .groups = 'drop'
  )
```

Within each site, there are different numbers of blocks: some sites have three sample blocks, others have four or five.
```{r}
toolik_plants %>% 
  group_by(Block) %>% 
  summarise(plot.n = n_distinct(Plot)) %>% 
  ungroup()
```

Within each block, there are eight smaller plots.
```{r}
unique(toolik_plants$Year)
```

How many species are represented in this data set?
```{r}
toolik_plants %>%
  select(Species) %>% 
  n_distinct()
```

There are 126 different species, but are they all actually species? It’s always a good idea to see what hides behind the numbers, so we can print the species to see what kind of species they are.
```{r}
toolik_plants %>% 
  select(Species) %>% 
  distinct() %>% 
  pull() %>%
  sort()
```

Some plant categories are in as just moss and lichen and they might be different species or more than one species, but for the purposes of the tutorial, we can count them as one species. There are other records that are definitely not species though: litter, bare (referring to bare ground), Woody cover, Tube, Hole, Vole trail, removed, vole turds, Mushrooms, Water, Caribou poop, Rocks, mushroom, caribou poop, animal litter, vole poop, Vole poop, Unk?.

You might wonder why people are recording vole poop. This relates to how the data were collected: each plot is 1m^2 and there are 100 points within it. When people survey the plots, they drop a pin from each point and then record everything that touches the pin, be it a plant, or vole poop!

The non-species records in the species column are a good opportunity for us to practice data manipulation. We will filter out the records we don’t need using the filter function from the dplyr package.
```{r}
toolik_plants <- toolik_plants %>% 
  filter(
    !Species %in% c('Woody cover', 'Tube', 'Hole', 
                    'Vole trail', 'removed', 'vole turds', 
                    'Mushrooms', 'Water', 
                    'Caribou poop', 'Rocks', 
                    'mushroom', 'caribou poop', 
                    'animal litter', 'vole poop', 
                    'Vole poop', 'Unk?')
  )
```

Let’s see how many species we have now:
```{r}
length(unique(toolik_plants$Species))
```

112 species! Next, we can calculate how many species were recorded in each plot in each survey year.
```{r}
toolik_plants <- toolik_plants %>% 
  group_by(Year, Site, Block, Plot) %>% 
  mutate(
    Richness = n_distinct(Species)
  ) %>% 
  ungroup()
```

To explore the data further, we can make a histogram of species richness.
```{r}
hist <- toolik_plants %>% 
  ggplot(aes(x = Richness))+
  geom_histogram()+
  theme_classic()
print(hist)
```

There are some other things we should think about. There are different types of numeric data here. For example, the years are whole numbers: we can’t have the year 2000.5.

The plant cover can be any value that is positive, it is therefore bounded at 0 and must be between 0 and 1. We can see this when we make a histogram of the data:
```{r}
hist2 <- toolik_plants %>% 
  ggplot(aes(x = Relative.Cover))+
  geom_histogram()+
  theme_classic()
print(hist2)
```

The plant cover data are skewed to the left, i.e., most of the records in the Relative.Cover column have small values. These distributions and characteristics of the data need to be taken into account when we design our model.

## 4. Thinking about our experimental design
In the Toolik dataset of plant cover, we have both spatial and temporal replication. The spatial replication is on three different levels: there are multiple sites, which have multiple blocks within them and each block has eight plots. The temporal replication refers to the different years in which plant cover was recorded: four years.

### Spatial autocorrelation
One of the assumptions of a model is that the data points are independent. In reality, that is very rarely the case. For example, plots that are closer to one another might be more similar, which may or may not be related to some of the drivers we’re testing, e.g. temperature.

### Temporal autocorrelation
Similarly, it’s possible that the data points in one year are not independent from those in the year before. For example, if a species was more abundant in the year 2000, that’s going to influence it’s abundance in 2001 as well.

## 5. Turn a question into a model
Let’s go back to our original question:

### Question 1: How has plant species richness changed over time at Toolik Lake?
What is our dependent and independent variable here? We could write out our base model in words:

Richness is a function of time.

In R, this turns into the code: richness ~ time.

Richness is our dependent (response) variable and time is our independent (predictor) variable ([see here for more details](https://en.wikipedia.org/wiki/Dependent_and_independent_variables)). This is our base model, but what other things do we need to account for? What would happen if we just modelled richness as a function of time without dealing with the other structure in our data? Let’s find out in the rest of the tutorial.

## 6. Learn about the different types of models
Before we get back to our dataset that we are designing a model for, let’s revisit some statistics basics.

**Here are some questions to consider.**

+ What is the difference between a continuous and a categorical variable in a linear model?
+ How many variables can you have in a model?
+ Is it better to have one model with five variables or one model per variable? When do we choose variables?
+ What is a fixed effect? What is a random effect?
+ What is the most important result from a model output?
+ Why does it matter which type of models we use?

## 7. General linear models
Model without any random effects:
```{r}
plant_m <- lm(Richness ~ I(Year-2007), data = toolik_plants)
summary(plant_m)
```

Notice how we have transformed the Year column - I(Year - 2007) means that the year 2008 will become Year 1 - then your model is estimating richness across the first, second, etc., year from your survey period. Otherwise, if we had kept the years just as 2008, 2009,…, the model would have estimated richness really far back into the past, starting from Year 1, Year 2… Year 1550 up until 2012. This would make the magnitude of the estimates we get wrong. You can experiment to see what happens if we just add in Year - suddenly the slope of species change goes in the hundreds!

**Assumptions made:**
+ The data are normally distributed.
+ The data points are independent of one another.
+ The relationship between the variables we are studying is actually linear.

And there are many more - you can check out [this useful website](https://r-statistics.co/Assumptions-of-Linear-Regression.html) for the full list with examples and how to check if those are assumptions are met later.

Do you think the assumptions of a general linear model are met for our questions and data? Probably not!

From the histograms, we can see that the data are not normally distributed, and furthermore, if we think about what the data are, they are integer counts (number of species), probably a bit skewed to the left, as most plots might not have a crazy amount of species. For these reasons, a Poisson distribution might be suitable, not a normal one. You can check out the Models and Distributions Coding Club tutorial for more about different data distributions.

We know that because of how the experimental design was set up (remember the Russian doll of plots within blocks within sites), the data points are not independent from one another. If we don’t account for the plot, block and site-level effects, we are completely ignoring the hierarchical structure of our data, which might then lead to wrong inferences based on the wrong model outputs.

### What is model convergence?
Model convergence is whether or not the model has worked, whether it has estimated your response variable (and random effects, see below) - basically whether the underlying mathematics have worked or have “broken” in some way. When we fit more complicated models, then we are pushing the limits of the underlying mathematics and things can go wrong, so it is important to check that your model did indeed work and that the estimates that you are making do make sense in the context of your raw data and the question you are asking/hypotheses that you are testing.

Checking model convergence can be done at different levels. With parametric models, good practice is to check the residual versus predicted plots. Using Bayesian approaches, there are a number of plots and statistics that can be assessed to determine model convergence. See below and in the Coding Club [MCMCglmm tutorial](https://ourcodingclub.github.io/tutorials/mcmcglmm/index.html). For an advanced discussion of model convergence, check out model [convergence in lme4](https://rdrr.io/cran/lme4/man/convergence.html).

For now, let’s check the residual versus predicted plot for our linear model. By using the ‘plot()’ function, we can plot the residuals versus fitted values, a Q-Q plot of standardised residuals, a scale-location plot (square roots of standardiaed residuals versus fitted values) and a plot of residuals versus leverage that adds bands corresponding to Cook’s distances of 0.5 and 1. Looking at these plots can help you identify any outliers that have huge leverage and confirm that your model has indeed run e.g. you want the data points on the Q-Q plot to follow the one-to-one line.
```{r}
plot(plant_m)
```

## 8. Hierarchical models using lme4
Now that we have explored the idea of a hierarchical model, let’s see how our analysis changes if we do or do not incorporate elements of the experimental design to the hierarchy of our model.

First, let’s model with only site as a random effect. This model does not incorporate the temporal replication in the data or the fact that there are plots within blocks within those sites.
```{r}
plant_m_plot <- lmer(Richness ~ I(Year-2007) + (1|Site), data = toolik_plants)
summary(plant_m_plot)
```

From the summary() outputs, you can see the effect sizes (under the column “Estimate” in the “Fixed effects” part of the summary), a key element of the model outputs. Effect sizes tell us about the strengths of the relationships we are testing. In this model, the “Year” variable has an effect of about -0.7 on “Richness”, which can be interpreted as an annual decrease of 0.7 species.

We are still not accounting for the different plots and blocks though, so let’s gradually add those and see how the results change.
```{r}
plant_m_plot2 <- lmer(Richness ~ I(Year-2007) + (1|Site/Block), data = toolik_plants)
summary(plant_m_plot2)
```

Have the estimates for the effect sizes changed?
```{r}
plant_m_plot3 <- lmer(Richness ~ I(Year-2007) + (1|Site/Block/Plot), data = toolik_plants)
summary(plant_m_plot3)
```

This final model answers our question about how plant species richness has changed over time, whilst also accounting for the hierarchical structure of the data.

Let’s check our model’s “fitted vs residuals” plot.
```{r}
plot(plant_m_plot3)
```

The points on this graph are evenly distributed on both sides of the horizontal line, which is a good sign that the model residuals do not violate the assumptions of the linear models.

Let's visualise the results using the `sjPlot` package!
```{r}
# set a clean theme for the graphs
set_theme(base = theme_bw() +
            theme(panel.grid.major.x = element_blank(),
                  panel.grid.minor.x = element_blank(),
                  panel.grid.minor.y = element_blank(),
                  panel.grid.major.y = element_blank(),
                  plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), units = , "cm")))

# Visualises random effects
re.effects <- plot_model(plant_m_plot3, type = 're', show.values = TRUE)
re.effects
```

Let's visualise our fixed effects:
```{r}
# To see the estimate for our fixed effect (default): Year
fe.effects <- plot_model(plant_m_plot3, show.values = TRUE)
fe.effects
```

Since we only have one fixed effect in this model (Year), the graph produced shows the estimate for that effect size (the point) and the confidence interval (line around the point).

Now, let’s look at the effect of mean temperature on richness. We will use the same hierarchical structure for site/block/plot random effects. We will also add year as random effect this time.
```{r}
plant_m_temp <- lmer(Richness ~ Mean.Temp + (1|Site/Block/Plot) + (1|Year), data = toolik_plants)
summary(plant_m_temp)
```

Let’s look at the fixed effect first this time:
```{r}
# Visualise the fixed effect
temp.fe.effects <- plot_model(plant_m_temp, show.values = TRUE)
temp.fe.effects
```

The very wide confidence interval this time suggest high uncertainty about the effect of temperature on richness.

And the random effects:
```{r}
# Visualise the tandom effect terms
temp.re.effects <- plot_model(plant_m_temp, type = 're', show.values = TRUE)
temp.re.effects
```

Assumptions made:

+ The data are normally distributed.
+ The data points are independent of one another.
+ The relationship between the variables we are studying is actually linear.
+ Plots represent the spatial replication and years represent the temporal replication in our data.

Assumptions not accounted for:

+ We have not accounted for spatial autocorrelation in the data - whether more closely located plots are more likely to show similar responses than farther away plots.
+ We have not accounted for temporal autocorrelation in the data - whether the influence of prior years of data are influencing the data in a given year.


























